---
title: "Impact of Vitamin C on Tooth Development"
author: "Jason R. Battles"
date: "September 22, 2016"
output: 
  pdf_document: 
    fig_height: 3.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# 1. Synopsis
This analysis will investigate the impact of Vitamin C on tooth development.  A study was performed in which 60 guinea pigs were given Vitamin C at various dosage levels (0.5, 1.0, and 2.0 mg/day) and by one of two delivery methods (Orange Juice and Ascorbic Acid).  The length of odontoblasts (cells responsible for tooth growth) were measured at the conclusion of the study.  

# 2. Prepare Environment & Data
Load the necessary libraries, ToothGrowth data, and make necessary modifications to the data.
```{r}
library(ggplot2)
library(datasets)
library(gridExtra)
data("ToothGrowth")
ToothGrowth$dose <- factor(ToothGrowth$dose) # Convert dose to factor
```

# 3. Data Summary & Exploratory Data Analysis
The data is a set of 60 observations. Each of these observations include variables for length of odontoblasts (teeth), dosage levels, and vitamin delivery method (OJ = Orange Juice, VC = Ascorbic Acid).  *R code for graph creation is included in the Appendix*

```{r, echo = FALSE}
ggplot(ToothGrowth,aes(dose,len)) + 
    geom_boxplot(aes(fill=supp)) + 
    scale_fill_brewer(palette="YlOrRd") +
    theme(legend.position=c(1,0.35),legend.justification=c(1,1)) + 
    labs(title="Scatter Plot of Tooth Length vs. Dosage Levels",x="Dosage Levels (mg)",y="Tooth Length")
```

Performing some basic exploratory data analysis using a factored scatterplot yields some preliminary observations:

1. Increased dosage levels appear to have a positive impact on tooth length.
2. Orange Juice (OJ) appears to have more positive impact on tooth length than Ascorbic Acid (AA) at lower dosage levels.

# 4. Compare Tooth Growth by Treatment
Further analysis using confidence intervals and hypothesis testing provides greater statistical insight into these preliminary observations

## 4.1 Tooth Growth by Dosage Level
The different dosage levels must first be prepared in order to create statistical comparisons
```{r}
dose_05_10 <- subset(ToothGrowth, dose %in% c(0.5, 1.0))
dose_05_20 <- subset(ToothGrowth, dose %in% c(0.5, 2.0))
dose_10_20 <- subset(ToothGrowth, dose %in% c(1.0, 2.0))
```

T tests are conducted on each of the dosage combinations to determine if the Null Hypothesis may be rejected for any of these.  The Null Hypothesis is that there is **no correlation** between the Dosage Levels and Tooth Growth.
```{r}
t.test(len ~ dose, paired = F, var.equal = F, data = dose_05_10)
t.test(len ~ dose, paired = F, var.equal = F, data = dose_05_20)
t.test(len ~ dose, paired = F, var.equal = F, data = dose_10_20)
```

The confidence intervals for each of these dosage levels allow the Null Hypothesis to be rejected.

* 0.5mg & 1.0mg Dosage Confidence Interval [-11.9838, -6.2762]  -> **Reject Null**
* 0.5mg & 2.0mg Dosage Confidence Interval [-18.1562, -12.8338] -> **Reject Null**
* 1.0mg & 2.0mg Dosage Confidence Interval [-8.9965, -3.7335]   -> **Reject Null**

**Increased Vitamin C dosage levels appear to have a positive effect on Tooth Growth.**

## 4.2 Tooth Growth by Delivery Method
Analyzing the data for statistical correlation between Tooth Growth and Delivery Type, the Null Hypothesis is tested.  The Null Hypothesis is that there is **no correlation** between Delivery Method and Tooth Growth.
```{r}
t.test(len ~ supp, paired = F, var.equal = F, data = ToothGrowth)
```
The 95% Confidence Interval is [-0.1710156, 7.5710156], which includes zero.  **Do not Reject the null hypothesis**

# Conclusions

1. There is statistical evidence to support that increased dosages of vitamin C have a positive effect on Tooth Growth.
2. There is not statistical evidence to support any finding that vitamin delivery method has any effect on Tooth Growth.

\newpage

# Appendix

## R Code for Initial Exploratory Data Analysis Scatter Plot
```{r, eval = FALSE}
ggplot(ToothGrowth,aes(dose,len)) + 
    geom_boxplot(aes(fill=supp)) + 
    scale_fill_brewer(palette="YlOrRd") +
    theme(legend.position=c(1,0.3),legend.justification=c(1,1)) + 
    labs(title="Scatter Plot of Tooth Length vs. Dosage Levels",x="Dosage Levels (mg)",y="Tooth Length")
```
